---
title: "Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Fluidigm)
```

# Preparations

## Requirements

Prior the installation of the Fluidigm package, you need to install plink, please install it according to the
Plink instructions on your system. The project webpage can be found here:

https://www.cog-genomics.org/plink2/

# Installation

## Cran (tba)
The `Fluidigm` package is hosted on Cran and can be installed via

```
install.packages("Fluidigm")
```

Stable release versions are usually even numbers like 0.2, 0.4, etc. 

## GitHub
The `Fluidigm` development platform is GitHub. The main-branch is usually similar to the Cran version. Also, published release versions on
Cran are similar to Cran. Hence, the same version as Cran can be installed from GitHub

```
library("devtools")
install_github("fischuu/Fluidigm")
```

Stable release versions are usually even numbers like 0.2, 0.4, etc.

However, the latest development version is hosted also on GitHub in the dev-branch and is labeled with uneven numbers 0.3.\*, 0.5.\* ,
where the \* indicates the running index of added features. These features are indicated in the CHANGELOG list and document the progress
of the package.

The latest development version of `Fluidigm` can be installed by running

```
library("devtools")
install_github("fischuu/Fluidigm@dev")
```

# Usage

The package is designed to run one function after another, namely


  1. `fluidigm2PLINK(...)`
  2. `estimateErrors(...)`
  3. `calculatePairwiseSimilarities(...)`
  4. `getPairwiseSimilarityLoci(...)`
  5. `similarityMatrix(...)`
  
and the user can either run them one by one or execute them all at once, using the convenient `fluidigmAnalysisWrapper(...)`-wrapper function.

## Example data and preparations

You can download the file example_data.csv from HERE e.g. to the folder `~/My_fluidigm_project` to follow the vignette steps.

Naturally, one first needs to load the `Fluidigm` library

```{r}
library("Fluidigm")
```

For convenience, we set first the working directory to avoid playing around with file paths

```{r, eval=FALSE}
setwd("~/My_fluidigm_project")
```
## Fluidigm to Plink

The first step of the package functionality is to create a ped/map-file pair from the csv output that is received from the Fluidigm machine.

The basic use of the function is like this
ß
```{r, eval=FALSE}
fluidigm2PLINK(file = "example_data.csv",
               map = "example_data.map",
               out = "new_data")
```

Here, the `file` parameter is the file that is received from the Fluidigm machine and the `map` option expects the location details
for the markers used in the Fluidigm, the `out` option defines the basename of the new generated files.

Other non-obligaroty input options are `out`, a string specifying the output file name. If left empty, the original basename of the input file will be used for the new generated output. Then the logical flag `plots` indicats whether additional figures for conversion should be plotted. Default is TRUE. With the logical flag `rearrange` it can be controled if the order of the created
ped/map filepair should follow the order given in the provided map file (option `TRUE`) or remain in the order as it is used in the Fluidigm file (option `FALSE`). The parameter `missing.geno` is a character string specifying how missing values should be coded. Default is "0 0". The logical option `fixNames` indicates whether whitespaces from sample names should be automatically removed. Default is TRUE. A file protection setting is the option `overwrite`. The logical indicates, wether the provided map sfile should be overwritten with the new map or not. By default, this is set to `FALSE` to avoid accidental file changes to the origianal file. If
it is set to `TRUE` without providing a new filename to `out`, the original file is overwritten, if it is set to `FALSE` the function requires an input to `out` to avoid overwriting of the original map file. As in all other functions, the functions talk-activity can be controlled with the numerical `verbosity` value. Set to a higher number for more details. Default is 1. Verbosity can be switched off with the logical flag `verbose`.

On case you want to overwrite the existing files and use the existing file names, the command would look like this

```{r, eval = FALSE}
fluidigm2PLINK(file = "example_data.csv",
               map = "example_data.map", 
               overwrite = TRUE)
```

### Output
The function generates besides the ped-file also a set of verbose plots (if the default option `plots = TRUE` is set). These are namely

1. `<filename>.additional_summary_stats.png`
2. `<filename>.call_rate_markers.png`
3. `<filename>.genotyping_success_samples.png`

The two outputs `<filename>.call_rate_markers.png` and `<filename>.genotyping_success_samples.png` show barplots to indicate the 
call rate of the genotypes and the genotyping success. The image`<filename>.additional_summary_stats.png` contains both of these,
plus some additional information like output filenames, number of samples and number of markers.

